Galois Slicing as Automatic Differentiation

### 2025-07-11 main/language.tex
Fix font.
Nicer formatting for CBN translation.

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes, where $\TWO = \{\top.\bot\}$:
  \begin{equation}\label{eqn:reverse-query-approx-1}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \mathbb{1} \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times \mathbb{1}
  \end{equation}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

@@

The monadic CBN translation is standard, and entirely determined by the translation on types, so we only
define the type translation $\cbn{-}$ here:
\begin{mathpar}
  \cbn{\rho} = \rho

  \cbn{\sigma \tySum \tau} = T(\cbn{\sigma}) \tySum T(\cbn{\tau})

  \cbn{\tyUnit} = \tyUnit

  \cbn{\sigma \tyProd \tau} = T(\cbn{\sigma}) \tyProd T(\cbn{\tau})

  \cbn{\sigma \tyFun \tau} = T(\cbn{\sigma}) \tyFun T(\cbn{\tau})

  \cbn{\tyList\;\tau} = \tyList\,(T(\cbn{\tau}))
\end{mathpar}
Thus, the CBN translation on types inserts a use of the monad $T(X) = \mathbb{A} \times X$ at the point ``just
underneath'' every type former. In this case, we are describing the type of data annotated at every
level. Note that our lists here are still ``strict'', an alternative approach would be to consider ``lazy''
lists that wrap the tail of every node in a $T$ as well.

>>>>

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes, where $\TWO = \{\top.\bot\}$:
  \begin{equation}\label{eqn:reverse-query-approx-1}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \mathbb{1} \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times \mathbb{1}
  \end{equation}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathit{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

@@

The monadic CBN translation is standard, and entirely determined by the translation on types, so we only
define the type translation $\cbn{-}$ here:
%\begin{mathpar}
%  \cbn{\rho} = \rho
%
%  \cbn{\sigma \tySum \tau} = T(\cbn{\sigma}) \tySum T(\cbn{\tau})
%
%  \cbn{\tyUnit} = \tyUnit
%
%  \cbn{\sigma \tyProd \tau} = T(\cbn{\sigma}) \tyProd T(\cbn{\tau})
%
%  \cbn{\sigma \tyFun \tau} = T(\cbn{\sigma}) \tyFun T(\cbn{\tau})
%
%  \cbn{\tyList\;\tau} = \tyList\,(T(\cbn{\tau}))
%\end{mathpar}
\[
\renewcommand{\arraystretch}{1.1}
\begin{array}{r@{\;=\;}l@{\qquad}r@{\;=\;}l@{\qquad}r@{\;=\;}l}
\cbn{\rho} & \rho &
\cbn{\sigma \tySum \tau} & T(\cbn{\sigma}) \tySum T(\cbn{\tau}) &
\cbn{\sigma \tyProd \tau} & T(\cbn{\sigma}) \tyProd T(\cbn{\tau}) \\
\cbn{\tyUnit} & \tyUnit &
\cbn{\sigma \tyFun \tau} & T(\cbn{\sigma}) \tyFun T(\cbn{\tau}) &
\cbn{\tyList\;\tau} & \tyList\,(T(\cbn{\tau}))
\end{array}
\]
Thus, the CBN translation on types inserts a use of the monad $T(X) = \mathbb{A} \times X$ at the point ``just
underneath'' every type former. In this case, we are describing the type of data annotated at every
level. Note that our lists here are still ``strict'', an alternative approach would be to consider ``lazy''
lists that wrap the tail of every node in a $T$ as well.
